Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-18 abstract.tex
edits upto Sec. 5

  Serializability is a well-understood correctness criterion that
  simplifies reasoning about the behavior of concurrent transactions
  by ensuring they are \emph{isolated} from each other while they
  execute.  However, enforcing serializable isolation comes at a steep
  cost in performance because it necessarily restricts opportunities
  to exploit concurrency even when such opportunities would not
  violate application-specific invariants. As a result, database
  systems in practice support, and often encourage, developers to
  implement transactions using weaker alternatives. These alternatives
  break the strong isolation guarantees offered by serializable
  transactions to permit greater concurrency. Unfortunately, the
  semantics of weak isolation is poorly understood, and usually
  explained only informally in terms of low-level implementation
  artifacts. Consequently, verifying high-level correctness properties
  in such environments remains a challenging problem.

  To address this issue, we present a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Recognizing that the proof burden
  necessary to use this logic may dissuade application developers, we
  also describe an inference procedure based on this foundation that
  ascertains the weakest isolation level that still guarantees the
  safety of high-level consistency assertions associated with such
  transactions.  The key to effective inference is the observation
  that weakly-isolated transactions can be viewed as functional
  (monadic) computations over an abstract database state, allowing us
  to treat their operations as state transformers over the database.
  This interpretation enables automated verification using
  off-the-shelf SMT solvers.

  Notably, our development is parametric over a transaction's specific
  isolation semantics, allowing it to be applicable over a range of
  concurrency control mechanisms.  Case studies and experiments on
  real-world applications (written in an embedded DSL in OCaml)
  demonstrate the utility of our approach, and provide strong evidence
  that automated verification of weakly-isolated transactions can be
  placed on the same formal footing as their strongly-isolated
  serializable counterparts.

>>>>

  Serializability is a well-understood correctness criterion that
  simplifies reasoning about the behavior of concurrent transactions
  by ensuring they are \emph{isolated} from each other while they
  execute.  However, enforcing serializable isolation comes at a steep
  cost in performance because it necessarily restricts opportunities
  to exploit concurrency even when such opportunities would not
  violate application-specific invariants. As a result, database
  systems in practice support, and often encourage, developers to
  implement transactions using weaker alternatives. These alternatives
  break the strong isolation guarantees offered by serializablity to
  permit greater concurrency. Unfortunately, the semantics of weak
  isolation is poorly understood, and usually explained only
  informally in terms of low-level implementation
  artifacts. Consequently, verifying high-level correctness properties
  in such environments remains a challenging problem.

  To address this issue, we present a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Recognizing that the proof burden
  necessary to use this logic may dissuade application developers, we
  also describe an inference procedure based on this foundation that
  ascertains the weakest isolation level that still guarantees the
  safety of high-level consistency invariants associated with such
  transactions.  The key to effective inference is the observation
  that weakly-isolated transactions can be viewed as functional
  (monadic) computations over an abstract database state, allowing us
  to treat their operations as state transformers over the database.
  This interpretation enables automated verification using
  off-the-shelf SMT solvers.

  Our development is parametric over a transaction's specific
  isolation semantics, allowing it to be applicable over a range of
  weak isolation mechanisms.  Case studies and experiments on
  real-world applications (written in an embedded DSL in OCaml)
  demonstrate the utility of our approach, and provide strong evidence
  that automated verification of weakly-isolated transactions can be
  placed on the same formal footing as their strongly-isolated
  serializable counterparts.

### 2017-10-18 motivation.tex
edits upto Sec. 5

\begin{figure}
\centering
\begin{ocaml}
let new_order (d_id, c_id, item_reqs) = atomically do
  dist <- SQL.select1 District (fun d -> d.d_id = d_id);
  let o_id = dist.d_next_o_id;
  SQL.update(* UPDATE *) District 
            (* SET *)(fun d -> {d with d_next_o_id =d_next_o_id + 1})
            (* WHERE *)(fun d -> d.d_id = d_id );
  SQL.insert(* INSERT INTO *) Order (* VALUES *){o_id=o_id;  
            o_d_id=d_id; o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
  foreach item_reqs @@ fun item_req -> do
    stk <- SQL.select1(* SELECT * FROM *) Stock 
              (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id 
                                  && s.s_d_id = d_id)(* LIMIT 1 *); 
    let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
                then stk.s_qty - item_req.ol_qty 
                else stk.s_qty - item_req.ol_qty + 91;
    SQL.update Stock (fun s -> {s with s_qty = s_qty'}) 
                     (fun s -> s.s_i_id = item_req.ol_i_id);
    SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
                           ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}

@@

TPC-C defines multiple invariants, called \emph{consistency
  conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
reflect the number of order lines under the order; this is the number
of \C{Order\_line} records whose \C{ol\_o\_id} field is the same as
\C{o.o\_id}.  In a sequential execution, it is easy to see how this
condition is preserved.  A new \C{Order} record is added with its
\C{o\_id} distinct from existing order ids, and its \C{o\_ol\_cnt} is
set to be equal to the size of the \C{item\_reqs} set. The \C{foreach}
loop runs once for each \C{item\_req}, adding a new \C{Order\_line}
record for each requested item, with its \C{ol\_o\_id} field set to
\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
records in the database (i.e., the number of records whose
\C{ol\_o\_id} field is equal to \C{o\_id}) is guaranteed to be equal
to the size of the \C{item\_req} set, which in turn is equal to the
\C{Order} record's \C{o\_ol\_cnt} field; these constraints ensure that
the transaction's consistency condition is preserved.

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in Fig.~\ref{fig:new_order_exec}.

@@

Notably, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes} - both concurrent transactions' writes are present
in the final state of the database. Indeed, program analyses that aim
to determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
can complicate high-level reasoning.  A possible alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts (i.e,, such a conflict occurs if
concurrentlyh executing transactions modify the same record) among
concurrent transactions when they commit. Write-write conflicts can be
eliminated in various ways, e.g., through conflict detection followed
by a rollback, or through exclusive locks, or a combination of both.
For instance, one possible implementation of SI, close to the one used
by PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

@@

As this discussion indicates, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction are not visible to another concurrently executing one
(call it $T$), during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witness
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from transaction variables to values, and let $\stl$ denote a
transaction-local log that maps the variables being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database does not change ($\Delta' =
\Delta$) during execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning sequentially.  When \C{new\_order}
finishes execution, however, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.  The axiomatization
precludes any execution in which there are concurrent updates to
shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
table), but does not prohibit interferences that write to different
tables, or write to different records in the same table.  We need to
reason about the safety of such interferences with respect to
\C{new\_order}'s consistency invariants.

@@

We approach the verification problem by first observing that a
relational database is a significantly simpler abstraction than shared
memory. There are no pointers, linked data structures, or aliasing.
Although a database essentially abstracts a mutable state, this state
is affected through a well-defined fixed number of interfaces (SQL
statements), each tagged with a logical formula describing what
records are accessed and updated.

This observation leads us away from thinking of a collection of
database transactions as a simple variant of a concurrent imperative
program.  Instead, we see value in viewing them as essentially
functional computations that manage database state monadically,
mirroring the structure of our DSL.  By doing so, we can formulate the
semantics of database operations as state transformers that explicitly
relate an operation's pre- and post-states, defining the semantics of
the corresponding transformer algorithmically, just like classical
predicate transformer semantics (e.g., weakest pre-condition or
strongest post-condition).  In our case, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
tuples, and a SQL statement as a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the tuples that match the search
criteria from $\Delta$ and computes the updated tuples that are to be
added to the local database. Thus, the state transformer for the update
statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
                s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}} 
           {\{\{s \with \C{s\_qty} = \C{k1}\}\}\\\hspace*{1.3in}}
           {\emptyset}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$. 
The transformer ($\T_I(\Delta,\delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \{\C{ol\_o\_id}=\C{k2};\, 
        \C{ol\_d\_id}=\C{k3};\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id};\, 
        \C{ol\_qty}=\C{item\_req.ol\_qty}\}
\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\Delta,\delta) =
\stl \cup F(\stg)$, where $F$ is a function that returns the set of
records added by the transformer to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
$\T_U$ and $\T_I$ shown above. The state transformation induced by the
loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
following composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stg) \cup \F_I(\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stg) \cup \F_I(\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
The advantage of inferring such transformers is that we can now make
use of a semantics-preserving translation from the domain of sets
equipped with $\bind$ to a decidable fragment of first-order logic,
allowing us to leverage SMT solvers for automated proofs without
having to infer potentially complex thread-local invariants or
intermediate assertions.  Sec.~\ref{sec:inference} describes this
translation. In the exposition thus far, we assumed $\Delta$ remains
invariant, which is clearly not the case when we admit concurrency.
Necessary concurrency extensions of the state transformer semantics to
deal with interference is also covered in Sec.~\ref{sec:inference}.
The following two sections, however, first focus on laying the
theoretical foundations for reasoning about weak isolation.

>>>>

\begin{figure}
\centering
\begin{ocaml}
let new_order (d_id, c_id, item_reqs) = atomically do
  dist <- SQL.select1 District (fun d -> d.d_id = d_id);
  let o_id = dist.d_next_o_id in
      SQL.update(* UPDATE *) District 
                (* SET *)(fun d -> {d with d_next_o_id =d_next_o_id + 1})
                (* WHERE *)(fun d -> d.d_id = d_id );
      SQL.insert(* INSERT INTO *) Order (* VALUES *){o_id=o_id;  
                o_d_id=d_id; o_c_id=c_id; o_ol_cnt=S.size item_reqs; };
      foreach item_reqs @@ fun item_req -> do
        stk <- SQL.select1(* SELECT * FROM *) Stock 
                  (* WHERE *)(fun s -> s.s_i_id = item_req.ol_i_id 
                                      && s.s_d_id = d_id)(* LIMIT 1 *); 
        let s_qty' = if stk.s_qty >= item_req.ol_qty + 10 
                    then stk.s_qty - item_req.ol_qty 
                    else stk.s_qty - item_req.ol_qty + 91 in
            SQL.update Stock (fun s -> {s with s_qty = s_qty'}) 
                             (fun s -> s.s_i_id = item_req.ol_i_id);
            SQL.insert Order_line {ol_o_id=o_id; ol_d_id=d_id; 
                                   ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}

@@

TPC-C defines multiple invariants, called \emph{consistency
  conditions}, over the state of the application in the database. One
such consistency condition is the requirement that for a given order
\C{o}, the \emph{order-line-count} field (\C{o.o\_ol\_cnt}) should
reflect the number of order lines under the order; this is the number
of \C{Order\_line} records whose \C{ol\_o\_id} field is the same as
\C{o.o\_id}.  In a sequential execution, it is easy to see how this
condition is preserved.  A new \C{Order} record is added with its
\C{o\_id} distinct from existing order ids, and its \C{o\_ol\_cnt} is
set to be equal to the size of the \C{item\_reqs} set. The \C{foreach}
loop runs once for each \C{item\_req}, adding a new \C{Order\_line}
record for each requested item, with its \C{ol\_o\_id} field set to
\C{o\_id}. Thus, at the end of the loop, the number of \C{Order\_line}
records in the database (i.e., the number of records whose
\C{ol\_o\_id} field is equal to \C{o\_id}) is guaranteed to be equal
to the size of the \C{item\_reqs} set, which in turn is equal to the
\C{Order} record's \C{o\_ol\_cnt} field; these constraints ensure that
the transaction's consistency condition is preserved.

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
default isolation level in 8 of the 18 databases studied
in~\cite{bailishotos}.  An executing RC transaction is isolated from
\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
allowed to witness the writes of concurrent transactions as soon as
they are committed. Thus, with two concurrent instances of the
\C{new\_order} transaction (call them $T_1$ and $T_2$), both
concurrently placing new orders for different customers under the same
district (\C{d\_id}), RC isolation allows the execution shown in
Fig.~\ref{fig:new_order_exec}.

@@

Clearly, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes} - both concurrent transactions' writes are present
in the final state of the database.  Program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
can complicate high-level reasoning.  A possible alternative would be
to utilize concurrent program verfication methods where the
implementation details. of weak isolation are interleaved within the
program, yielding a (more-or-less) conventional concurrent program;
but, considering the size and complexity of real-world transaction
systems, this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts (i.e., such a conflict occurs if
concurrently executing transactions modify the same record) among
concurrent transactions when they commit. Write-write conflicts can be
eliminated in various ways, e.g., through conflict detection followed
by a rollback, or through exclusive locks, or a combination of both.
For instance, one possible implementation of SI, close to the one used
by PostgreSQL~\cite{postgresiso}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

@@

As this discussion hints, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction (call it $T$) are not visible to another concurrently executing one
during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witnesses
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from database locations to values, and let $\stl$ denote a
transaction-local log that maps the locations being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\mathit{dom}(delta)).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database does not change ($\Delta' =
\Delta$) during execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning sequentially.  When \C{new\_order}
finishes execution, however, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The axiomatization
precludes any execution in which there are concurrent updates to
shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
table), but does not prohibit interferences that write to different
tables, or write to different records in the same table.  We need to
reason about the safety of such interferences with respect to
\C{new\_order}'s consistency invariants.

@@

We approach the verification problem by first observing that a typical
relational database is a significantly simpler abstraction than shared
memory.  Its primary data structure is a table, with no primitive
support for pointers, linked data structures, or aliasing.  Although a
database essentially abstracts a mutable state, this state is managed
through a well-defined fixed number of interfaces (SQL statements),
each tagged with a logical formula describing what records are
accessed and updated.

This observation leads us away from thinking of a collection of
database transactions as a simple variant of a concurrent imperative
program.  Instead, we see value in viewing them as essentially
functional computations that manage database state monadically,
mirroring the structure of our DSL.  By doing so, we can formulate the
semantics of database operations as state transformers that explicitly
relate an operation's pre- and post-states, defining the semantics of
the corresponding transformer algorithmically, just like classical
predicate transformer semantics (e.g., weakest pre-condition or
strongest post-condition).  In our case, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
records, and a SQL statement as a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

@@

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1},\C{k2}, and \C{k3}).  The body of
the loop executes a SQL update followed by an insert.  Recall that a
transaction reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these
updates. An update statement filters the records that match the search
criteria from $\Delta$ and computes the updated records that are to be
added to the local database. Thus, the state transformer for the
update statement (call it $\T_U$) is the following function on
sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \stg \bind(\lambda s. 
      \itel{{\sf table}(s) = \C{Stock} \conj 
        s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\\hspace*{1.3in}}
           {\{ \langle s\ {\sf with}\  s_\C{qty} = \C{k1}\rangle \}\\\hspace*{1.3in}}
           {\emptyset})
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
% Observe that
% $\T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
% F_U(\Delta)$.
Here, the set bind operator extracts record elements ($\C{s}$) from
the database, checks the precondition of the update action, and if
satisfied, constructs a new set containing a single record that is
identical to $\C{s}$ except that it binds field $\C{s_{qty}}$ to value
$\C{k_1}$.  This new set is added (via set union) to the existing
local database state $\delta$.\footnote{For now, assume that the
  record being added to the local database has not previously been
  updated.}

The transformer ($\T_I(\delta,\Delta)$) for the subsequent \C{insert}
statement can be similarly constructed:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup
       \{\langle\C{ol\_o\_id}=\C{k2},
                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}

\end{array}
\end{smathpar}
Observe that both transformers are of the form $T(\delta,\Delta) =
\stl \cup F(\stl,\stg)$, where $F$ is a function that returns the set
of records added  to the transaction-local database
($\stl$). Let $\F_U$ and $\F_I$ be the corresponding functions for
$\T_U$ and $\T_I$ shown above. The state transformation induced by the
loop body in Fig.~\ref{fig:new_order_code} can be expressed as the
following composition of $\F_U$ and $\F_I$:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \F_U(\stl,\stg) \cup \F_I(\stl,\stg)
\end{array}
\end{smathpar}
The transformer for the loop itself can now be computed to be:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Observe that the structure of the transformer mirrors the structure of
the program itself. In particular, SQL statements become set
operations, and the \C{DB} monad's $\bind$ becoms set monadic $\bind$.
As we demonstrate, the advantage of inferring such transformers is
that we can now make use of a semantics-preserving translation from
the domain of sets equipped with $\bind$ to a decidable fragment of
first-order logic, allowing us to leverage SMT solvers for automated
proofs without having to infer potentially complex thread-local
invariants or intermediate assertions.  Sec.~\ref{sec:inference}
describes this translation. In the exposition thus far, we assumed
$\Delta$ remains invariant, which is clearly not the case when we
admit concurrency.  Necessary concurrency extensions of the state
transformer semantics to deal with interference is also covered in
Sec.~\ref{sec:inference}.  Before presenting the transformer
semantics, we first focus our attention in the following two sections
on the theoretical foundations for weak isolation, upon which this
semantics is based.
